On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2

### 2023-09-13 manuscript/pnas_revision.tex
finished rev 1 other than sim

The left plot of Figure \ref{fig:post} presents posterior densities
for the fixed-effect coefficients from regression \eqref{eq:mixed}.
The air traffic proximity, intracontinental proximity and Hubei
asymmetry coefficients have posterior means and 95\% credible
intervals of 0.76 (0.50, 1.03), 0.04 (-0.04, 0.12) and 0.27 (-0.39,
0.78), respectively. While these results suggest a positive
association between each predictor matrix and the infinitesimal
generator matrix $\QQ$, the only predictor with a statistically
significant association is air traffic proximity.  This result agrees
with a previous phylogeographic analysis of the global spread of
influenza \cite{holbrook2021massive}.  The right plot of Figure
\ref{fig:post} presents the posterior mean for generator $\QQ$.  As
one may expect, the matrix looks similar to that of the air traffic
predictor matrix in Figure \ref{fig:FE}, but one may also see the
influence of the Hubei asymmetry in, e.g., the squares corresponding
to travel between Guangdong and Hubei provinces.  Finally, we randomly
generate regions of unobserved ancestors from their posterior
predictive distributions every 100-thousandth MCMC iteration.
Figure \ref{fig:tree} projects the empirical posterior predictive mode
of these locations onto the phylogenetic tree $\phylogeny$.

\begin{figure*}[!t]
	\centering
	\includegraphics[width=0.7\textwidth]{summary_tree.pdf}
	\vspace{-1em}
	\caption{Posterior predictive modes for (unobserved) ancestral
          locations color a phylogenetic tree that describes the
          shared evolutionary history of 285 SARS-CoV-2
          samples.}\label{fig:tree}
\end{figure*}

>>>>

The left plot of Figure \ref{fig:post} presents posterior densities
for the fixed-effect coefficients from regression \eqref{eq:mixed}.
The air traffic proximity, intracontinental proximity and Hubei
asymmetry coefficients have posterior means and 95\% credible
intervals of 0.76 (0.50, 1.03), 0.04 (-0.04, 0.12) and 0.27 (-0.39,
0.78), respectively. While these results suggest a positive
association between each predictor matrix and the infinitesimal
generator matrix $\QQ$, the only predictor with a statistically
significant association is air traffic proximity.  This result agrees
with a previous phylogeographic analysis of the global spread of
influenza \cite{holbrook2021massive}.  The right plot of Figure
\ref{fig:post} presents the posterior mean for generator $\QQ$.  As
one may expect, the matrix looks similar to that of the air traffic
predictor matrix in Figure \ref{fig:FE}, but one may also see the
influence of the Hubei asymmetry in, e.g., the squares corresponding
to travel between Guangdong and Hubei provinces.  Finally, we randomly
generate regions of unobserved ancestors from their posterior
predictive distributions every 100-thousandth MCMC iteration.
After collapsing regions into 8 major blocks, Figure \ref{fig:tree} projects the empirical posterior predictive mode of these blocks onto the phylogenetic tree $\phylogeny$.  The general pattern looks similar to that of Figure 1 from \cite{lemey2020accommodating}, although the geographic blocking scheme differs slightly.

\begin{figure*}[!t]
	\centering
	\includegraphics[width=0.7\textwidth]{summary2.tree.pdf}
	\vspace{-0.5em}
	\caption{Posterior predictive modes for (unobserved) ancestral
          locations color a phylogenetic tree that describes the
          shared evolutionary history of 285 SARS-CoV-2
          samples.}\label{fig:tree}
\end{figure*}
